Topic categorization in Portuguese news articles\\ \thanks{Fundação para a Ciência e Tecnologia}

### 2018-01-03 article.tex
references

\begin{abstract} This document is a model and instructions for \LaTeX.
This and the IEEEtran.cls file define the components of your paper
[title, text, heads, etc.]. *CRITICAL: Do Not Use Symbols, Special
Characters, Footnotes, or Math in Paper Title or Abstract.
\end{abstract}

@@

\section{Introduction}
Online news articles first appeared as reprints from traditional
newspapers; nowadays, however, they represent now the primary source
of news for some segments of the population, both in developed and
developing countries (whether consumed directly in the newspaper
website, or indirectly e.g. through a social media application or a
feed catcher).

Unofficially know as \emph{the fourth branch of government}, the press
plays a vital role within our society, keeping us informed about
the current state of affairs (at a local and global scale) and acting
as a watchdog for the other three branches (legislative, executive and
judicial). The (lack of) freedom of press and access to the news in a
given country is even often considered an indicator of a lack of
democracy\cite{goode2009social}.

As such, improving the ways citizens can access the information (view
it, query it and search it) contained in news articles has the
potential to contribute for a more informed and, ultimately, a
better society.

On the other hand, the last decades have witnessed a fast increase on
the rate of publication of digital text documents. Traditional
document types, such as news articles, scientific papers or books are
now published online together with new formats, such as blog posts or
tweets, each having thousands or millions of new documents published
each day.

And it was not only the publication step which has moved to the
digital world; in fact, most often nowadays the whole document
lifecycle happens digitally, with virtual tools available for
researching, writing, styling, publishing and sharing.

Having the entire workflow happening within the digital world presents
some opportunities when compared to the traditional process. In
particular, due to the current processing power commonly available,
tasks related to the manipulation of the information contained within
these documents (searching, compiling, annotating, sharing, \dots) can
now be performed automatically and targetting a large amount of
articles.

@@

An example of a feature which improves information access is the
categorization of news articles by the topic (or topics) of its
content. The presence of such a categorization may influence the way
the information is stored, organized, displayed and queried.

@@

\begin{itemize}
    \item the term length was between 3 and 30 characters (to discard
        things like URLs or badly tokenized sentences)
    \item the term appeared in the document at least twice
    \item the term appeared at least in 10 documents
\end{itemize}
The algorithm used to weigh the terms in the matrix was the
\textit{tf-idf}, and resulted in 3234 terms.

>>>>

\begin{abstract} 
Categorizing news articles according to their contents allows to
decrease the information entropy in a world where the rate of
publication of digital text documents is increasing fast. In this
article we describe ongoing work which aims to evaluate the
feasability of implementing a classifier which is lightweight
enough to be used in real time on the client side of a web
application.  More specifially, we gathered a corpus of Portuguese
news and used it to train and evaluate several classification
algorithms. We analyze the results obtained in terms of the
classifiers error rate, traning time and memory footprint.
\end{abstract}

@@

\section{Introduction}
Online news articles first appeared as reprints from traditional
newspapers; nowadays, however, they represent now the primary source
of news for some segments of the population, both in developed and
developing countries (whether consumed directly in the newspaper
website, or indirectly e.g. through a social media application or a
feed catcher)\cite{greer2006evolution,boczkowski2005digitizing,chyi1999access}.

Unofficially know as \emph{the fourth branch of government}, the press
plays a vital role within our society, keeping us informed about
the current state of affairs (at a local and global scale) and acting
as a watchdog for the other three branches (legislative, executive and
judicial). The (lack of) freedom of press and access to the news in a
given country is even often considered an indicator of a lack of
democracy\cite{goode2009social,house2009freedom}.

As such, improving the ways citizens can access the information (view
it, query it and search it) contained in news articles has the
potential to contribute for a more informed and, ultimately, a
better society\cite{bollinger1988tolerant}.

On the other hand, the last decades have witnessed a fast increase on
the rate of publication of digital text documents. Traditional
document types, such as news articles, scientific papers or books are
now published online together with new formats, such as blog posts or
tweets, each having thousands or millions of new documents published
each day\cite{hilbert2011world,allan2006online}.

And it was not only the publication step which has moved to the
digital world; in fact, most often nowadays the whole document
lifecycle happens digitally, with virtual tools available for
researching, writing, styling, publishing and sharing\cite{williams2009personal}.

Having the entire workflow happening within the digital world presents
some opportunities when compared to the traditional
process\cite{o1997comparison}. In particular, due to the current
processing power commonly available, tasks related to the manipulation
of the information contained within these documents (searching,
compiling, annotating, sharing, \dots) can now be performed
automatically and targetting a large amount of articles.

@@

An example of a feature which improves information access is the
categorization of news articles by the topic (or topics) of its
content\cite{kim2006extracting}. The presence of such a categorization
may influence the way the information is stored, organized, displayed
and queried.

@@

\begin{itemize}
    \item the term length was between 3 and 30 characters (to discard
        things like URLs or badly tokenized sentences)
    \item the term appeared in the document at least twice
    \item the term appeared at least in 10 documents
\end{itemize}
The algorithm used to weigh the terms in the matrix was the
\textit{tf-idf}\cite{christopher2008introduction}, and resulted in 3234 terms.
